Reproducible Research in R: A Tutorial on how to Do the Same Thing More Than Once

### 2021-11-19 manuscript.Rmd

---
title: "Reproducible Research in R: A Tutorial on how to Do the Same Thing More Than Once"
author:
  - name: Aaron Peikert
    affil: 1, 2, *
    orcid: 0000-0001-7813-818X
  - name: Caspar J. van Lissa
    affil: 3, 4
    orcid: 0000-0002-0808-5024
  - name: Andreas M. Brandmaier
    affil: 1, 5, 6
    orcid: 0000-0001-8765-6982
affiliation:
  - num: 1
    address: |
      Center for Lifespan Psychology---Max Planck Institute for Human Development, Lentzeallee 94, 14195 Berlin, Germany
  - num: 2
    address: |
      Humboldt-Universität zu Berlin, Berlin, Germany
  - num: 3
    address: |
      Department of Methodology & Statistics---Utrecht University, Faculty of Social and Behavioral Sciences, Utrecht, Netherlands
  - num: 4
    address: |
      Open Science Community Utrecht, Utrecht, Netherlands
  - num: 5
    address: |
      Max Planck UCL Centre for Computational Psychiatry and Ageing Research
      Berlin, Germany, and London, UK
  - num: 6
    address: |
      MSB Medical School Berlin,
      Berlin, Germany
# firstnote to eighthnote
correspondence: |
  peikert@mpib-berlin.mpg.de
journal: psych
type: tutorial
status: submit
bibliography: temp.bib
simplesummary: |
  Reproducibility has long been considered integral to the scientific method.
  An analysis is considered reproducible if an independent person can obtain the same results from the same data.
  Until recently, detailed descriptions of methods and analyses were the primary instrument for ensuring scientific reproducibility.
  Technological advancements now enable scientists to achieve a more comprehensive standard
  that allows anyone to access a digital research repository and reproduce all computational steps from raw data to final report, including all relevant statistical analyses, with a single command.
  This method has far-reaching implications for scientific archiving,
  reproducibility and replication,
  scientific productivity,
  and the credibility and reliability of scientific knowledge.
  One obstacle to the widespread use of this method is that the underlying tools
  are complex and not part of most researchers' basic training.
  This paper introduces `repro`, an R package that guides researchers through installation and use of the tools required to make a research project reproducible.
  We also suggest using the proposed workflow for the preregistration of study plans as reproducible computer code (preregistration as code; PAC).
  Since computer code represents the planned analyses exactly as they will be executed, it is more precise than natural language descriptions. 
  PAC circumvents the shortcomings of ambiguous preregistrations that may result in undisclosed researcher degrees of freedom when adjustments to the planned analyses are necessary.
  Reproducibility, facilitated by automation, has a wide range of applications and could potentially accelerate scientific progress.
abstract: "`r tryCatch(trimws(readr::read_file(here::here('abstract.Rmd'))))`"
keywords: |
  open science; computational reproducibility; preregistration; R; R Markdown; Make; GitHub; Docker
acknowledgement: |
  We would like to thank Maximilian Stefan Ernst for his contributions to the code for the simulation study. We are grateful to Julia Delius for her helpful assistance in language and style editing. The R package `repro` was developed as part of the [first author's master thesis](https://aaronpeikert.github.io/repro-thesis/) at the Humboldt-Universität zu Berlin.
authorcontributions: |
  Aaron Peikert took the lead in writing and provided the initial draft of the manuscript.
  Andreas Brandmaier and Caspar J. Van Lissa contributed further ideas, critical feedback, and revisions of the original manuscript.

@@

With this paper, we pursue two aims.
First, we want to introduce researchers to a notion of automated reproducibility that requires no manual steps, apart from the initial setup of the software environment. 
Secondly, we discuss the implications of automated reproducibility for changing the general approach to research.
With regard to the first goal, we discuss how to address four common threats to reproducibility, using tools originating from software engineering [see @Peikert2019 for details], and present a tutorial on how to employ these tools to achieve automated reproducibility.
A single tutorial cannot comprehensively introduce the reader to the individual tools, but this tutorial is intended to help readers get started with a basic workflow.
The tutorial is aimed at researchers who regularly write code to analyze their data and are willing to make relevant code, data, and materials available, either publicly or on request.
Ideally, the reader has already created a dynamic document at some point in time (e.g., with R Markdown or Jupyter) and used some form of version control (e.g., Git). 
The R package `repro` supports researchers in setting up the required software and in adopting this workflow.
We present automated reproducibility as a best practice; a goal that is not always fully achieved due to limited resources, technical restrictions, or practical considerations, but is worth striving for nonetheless.

In pursuit of the second aim, we present a strictly reproducible and unambiguous form of preregistration [@NosekRevolution2018] that builds upon implementing this reproducible workflow, the so-called *Preregistration as Code* (PAC).
PAC involves preregistering the intended analysis code and the major part of the final scientific report as a dynamic document, including typical sections like introduction, methods, and results.
The resulting dynamic document closely resembles the final manuscript but uses simulated data to generate placeholder results (e.g., figures, tables, and statistics).
Simulated data allows authors to write their intended analysis script and preregister it<!-- CJ: Aaron: I don't think the key point is that users preregister the presentation of their results (tables, figures), but rather that they preregister the analysis!!!-->.
Once the empirical data are available, these replace the simulated data; the results are then updated automatically, and the discussion can be written to finalize the report.

@@

`Git` tracks changes to all project-related files (e.g., materials, data, and code) over time.
At any stage, individual files or the entire project can be compared to, or reverted to, an earlier version.
Moreover, contributions (e.g., from collaborators) can be compared to, and incorporated in the master version of the project. 
Version control thus reduces the risk of losing work and facilitates collaboration.
`Git` is built around snapshots that represent the project state at a given point in time.
These snapshots are called *commits* and work like a "save" action.
Ideally, each commit has a message that succinctly describes these changes.
It is good practice to make commits for concrete milestones (e.g., "Commented on Introduction," "Added SES as a covariate," "Address Reviewer 2's comment 3").
This makes it easier to revert specific changes than when multiple milestones are joined in one commit, e.g., "Changes made on 19/07/2021".
Each commit refers back to its ancestor, and all commits are thus linked in a timeline.
The entirety of commits (i.e., the version-controlled project) is called a repository.
In `Git`, specific snapshots of a repository can be tagged, such that the user can clearly label which version of the project was used to create a preregistration, preprint, or final version of the manuscript as accepted by a journal.
`Git` has additional features beyond basic version control, such as "branches" (parallel versions of a project that can later be merged again) to facilitate simultaneous collaboration.
@vuorreCuratingResearchAssets2018 provide a more extensive treatment of how `Git` functions and how to use `Git` for research.
@bryanExcuseMeYou2018 provides additional information on how to track `R Markdown` documents.
Collaborating via `Git` is facilitated by uploading the repository to a cloud-based service.<!-- CJ This is not true-->
We recommend GitHub as a host for `Git` repositories because of its popularity among R users.
GitHub has many tools that facilitate working with `Git` - in particular project management and collaboration - but these are not central to achieving reproducibility.

@@

Dynamic document generation integrates both steps.
Through dynamic document generation, code becomes an integral, although usually hidden, part of the manuscript, complementing the verbal description and allowing interested readers to gain a deeper understanding of the contents [@knuthCWEBSystemStructured; @claerboutElectronicDocumentsGive1992].
`R Markdown` uses `Markdown` for text formatting and R (or other programming languages) for writing the statistical analysis.
`Markdown` is a lightweight text format in plain text with a minimal set of reserved symbols for formatting instructions.
This way, `Markdown` does not need any specialized software for editing.
It is fool-proof <!-- CJ: I don't understand what this means-->(unlike, for example, `LaTeX` [@lamportLATEXDocumentPreparation1994]), works well with version control systems, and can be exported to various document formats, such as HTML websites, a Microsoft Word document, a typeset PDF file (for example, via LaTeX journal templates), or a Powerpoint presentation.
`Markdown` can be used for all sorts of academic documents, ranging from simple sketches of ideas to scientific manuscripts [@R-rticles] and presentations [@revealjs], or even résumés [@vitae].
`R Markdown` extends regular `Markdown` by allowing users to include R code chunks (in fact, arbitrary computer code [@xieMarkdownCookbook2020, Chapter 15, Chapter 15, Other Languages]) into a `Markdown` document.
Upon rendering the document, the code blocks are executed, and their output is dynamically inserted into the document.
This allows the creation of (conditionally) formatted text, statistical results, and figures that are guaranteed to be up-to-date because they are created anew every time the document is rendered to its output format (e.g., presentation slides or a journal article).
@xieMarkdownCookbook2020 provides an extensive yet practical introduction to most features of `R Markdown`.

@@

*Dependency tracking* helps automatically resolve dependencies between project files.
In essence, researchers provide a collection of *computational recipes*.
A computational recipe describes how inputs are processed to deterministically create a specific output in a way that is automatically executable.
The concept of computational recipes is central to our understanding of reproducibility because it enables a unified way to reproduce a project automatically.
Similar to a collection of cooking recipes, we can have multiple products (_targets_) with different ingredients (_requirements_) and different steps of preparation (_recipes_).
In the context of scientific data analysis, targets are typically the final scientific report (e.g., the one to be submitted to a journal) and possibly intermediate results (such as preprocessed data files, simulation results, and analysis results).
One noteworthy property of computational recipes is that they require a textual representation and the ability of a program to make sense of this textual representation.<!-- CJ I don't understand, and is it necessary to explain this?-->
A workflow that involves renaming variable names by hand in a graphical spreadsheet application, for example, is therefore incompatible with automated reproducibility.
Another property of a computational recipe is that the same inputs should always result in the same outputs.
For most computer code (given the same software is used), this property is fulfilled.
However, one noteworthy exception is the generation of pseudo-random numbers.
Whenever random numbers are used in a computation, it is only reproducible if the random number generator generates the same numbers.
To ensure identical random numbers, users may fix the state of the random number generated with a so-called seed (e.g. `set.seed()` in R), but they also need to guarantee that the pseudo-random number generator is unchanged [see @Peikert2019].

@@

A version-controlled dynamic document with dependency tracking still relies on external software (even if just to be rendered to its final output format).
Troubleshooting issues specific to a particular programming language or dependent tool typically require considerable expertise and threaten reproducibility.
*Software management* refers to the act of providing records of, or access to, all software packages and system libraries a project depends on.
One comprehensive approach to software management is containerization.
The central idea is that by
"[..] packaging the key elements of the computational environment needed to run the desired software [makes] the software much easier to use, and the results easier to reproduce [...]" [@silverSoftwareSimplified2017, p. 174].

>>>>

---
title: "Reproducible Research in R: A Tutorial on how to Do the Same Thing More Than Once"
author:
  - name: Aaron Peikert
    affil: 1, 2, *
    orcid: 0000-0001-7813-818X
  - name: Caspar J. van Lissa
    affil: 3, 4
    orcid: 0000-0002-0808-5024
  - name: Andreas M. Brandmaier
    affil: 1, 5, 6
    orcid: 0000-0001-8765-6982
affiliation:
  - num: 1
    address: |
      Center for Lifespan Psychology---Max Planck Institute for Human Development, Lentzeallee 94, 14195 Berlin, Germany
  - num: 2
    address: |
      Humboldt-Universität zu Berlin, Berlin, Germany
  - num: 3
    address: |
      Department of Methodology & Statistics---Utrecht University, Faculty of Social and Behavioral Sciences, Utrecht, Netherlands
  - num: 4
    address: |
      Open Science Community Utrecht, Utrecht, Netherlands
  - num: 5
    address: |
      Max Planck UCL Centre for Computational Psychiatry and Ageing Research
      Berlin, Germany, and London, UK
  - num: 6
    address: |
      MSB Medical School Berlin,
      Berlin, Germany
# firstnote to eighthnote
correspondence: |
  peikert@mpib-berlin.mpg.de
journal: psych
type: tutorial
status: submit
bibliography: temp.bib
simplesummary: |
  Reproducibility has long been considered integral to the scientific method.
  An analysis is considered reproducible if an independent person can obtain the same results from the same data.
  Until recently, detailed descriptions of methods and analyses were the primary instrument for ensuring scientific reproducibility.
  Technological advancements now enable scientists to achieve a more comprehensive standard that allows anyone to access a digital research repository and reproduce all computational steps from raw data to final report, including all relevant statistical analyses, with a single command.
  This method has far-reaching implications for scientific archiving,
  reproducibility and replication,
  scientific productivity,
  and the credibility and reliability of scientific knowledge.
  One obstacle to the widespread use of this method is that the underlying tools
  are complex and not part of most researchers' basic training.
  This paper introduces `repro`, an R package that guides researchers through installation and use of the tools required to make a research project reproducible.
  We also suggest using the proposed workflow for the preregistration of study plans as reproducible computer code (preregistration as code; PAC).
  Since computer code represents the planned analyses exactly as they will be executed, it is more precise than natural language descriptions. 
  PAC circumvents the shortcomings of ambiguous preregistrations that may result in undisclosed use of researcher degrees of freedom.
  Reproducibility, facilitated by automation, has a wide range of applications and could potentially accelerate scientific progress.
abstract: "`r tryCatch(trimws(readr::read_file(here::here('abstract.Rmd'))))`"
keywords: |
  open science; computational reproducibility; preregistration; R; R Markdown; Make; GitHub; Docker
acknowledgement: |
  We would like to thank Maximilian Stefan Ernst for his contributions to the code for the simulation study. We are grateful to Julia Delius for her helpful assistance in language and style editing. The R package `repro` was developed as part of the [first author's master thesis](https://aaronpeikert.github.io/repro-thesis/) at the Humboldt-Universität zu Berlin.
authorcontributions: |
  Aaron Peikert took the lead in writing and provided the initial draft of the manuscript.
  Andreas Brandmaier and Caspar J. Van Lissa contributed further ideas, critical feedback, and revisions of the original manuscript.

@@

With this paper, we pursue two aims.
First, we want to introduce researchers to a notion of automated reproducibility that requires no manual steps, apart from the initial setup of the software environment. 
Secondly, we discuss the implications of automated reproducibility for changing the general approach to research.
With regard to the first goal, we discuss how to address four common threats to reproducibility, using tools originating from software engineering [see @Peikert2019 for details], and present a tutorial on how to employ these tools to achieve automated reproducibility.
A single tutorial cannot comprehensively introduce the reader to the detail of individual tools, but this tutorial is intended to help readers get started with a basic workflow.
The tutorial is aimed at researchers who regularly write code to analyze their data and are willing to make relevant code, data, and materials available, either publicly or on request.
Ideally, the reader has already created a dynamic document at some point in time (e.g., with R Markdown or Jupyter) and used some form of version control (e.g., Git). 
The R package `repro` supports researchers in setting up the required software and in adopting this workflow.
We present automated reproducibility as a best practice; a goal that is not always fully achieved due to limited resources, technical restrictions, or practical considerations, but is worth striving for nonetheless.

In pursuit of the second aim, we present a strictly reproducible and unambiguous form of preregistration [@NosekRevolution2018] that builds upon implementing this reproducible workflow, the so-called *Preregistration as Code* (PAC).
PAC involves preregistering the intended analysis code and the major part of the final scientific report as a dynamic document, including typical sections like introduction, methods, and results.
The resulting dynamic document closely resembles the final manuscript but uses simulated data to generate placeholder results (e.g., figures, tables, and statistics).
Simulated data serve two functions, they allow to test the code for the planned analyses and for preregistering the exact presentation of the results.
Once the empirical data are available, these replace the simulated data; the results are then updated automatically, and the discussion can be written to finalize the report.

@@

`Git` tracks changes to all project-related files (e.g., materials, data, and code) over time.
At any stage, individual files or the entire project can be compared to, or reverted to, an earlier version.
Moreover, contributions (e.g., from collaborators) can be compared to, and incorporated in the main version of the project. 
Version control thus reduces the risk of losing work and facilitates collaboration.
`Git` is built around snapshots that represent the project state at a given point in time.
These snapshots are called *commits* and work like a "save" action.
Ideally, each commit has a message that succinctly describes these changes.
It is good practice to make commits for concrete milestones (e.g., "Commented on Introduction," "Added SES as a covariate," "Address Reviewer 2's comment 3").
This makes it easier to revert specific changes than when multiple milestones are joined in one commit, e.g., "Changes made on 19/07/2021".
Each commit refers back to its ancestor, and all commits are thus linked in a timeline.
The entirety of commits (i.e., the version-controlled project) is called a repository.
In `Git`, specific snapshots of a repository can be tagged, such that the user can clearly label which version of the project was used to create a preregistration, preprint, or final version of the manuscript as accepted by a journal.
`Git` has additional features beyond basic version control, such as "branches" (parallel versions of a project that can later be merged again) to facilitate simultaneous collaboration.
@vuorreCuratingResearchAssets2018 provide a more extensive treatment of how `Git` functions and how to use `Git` for research.
@bryanExcuseMeYou2018 provides additional information on how to track `R Markdown` documents.
Collaborating via `Git` is facilitated by uploading the repository to a cloud-based service.
We recommend GitHub as a host for `Git` repositories because of its popularity among R users.
GitHub has many tools that facilitate working with `Git` --- in particular project management and collaboration --- but these are not central to achieving reproducibility.

@@

Dynamic document generation integrates both steps.
Through dynamic document generation, code becomes an integral, although usually hidden, part of the manuscript, complementing the verbal description and allowing interested readers to gain a deeper understanding of the contents [@knuthCWEBSystemStructured; @claerboutElectronicDocumentsGive1992].
`R Markdown` uses `Markdown` for text formatting and R (or other programming languages) for writing the statistical analysis.
`Markdown` is a lightweight text format in plain text with a minimal set of reserved symbols for formatting instructions.
This way, `Markdown` does not need any specialized software for editing.
It is userfriendly (unlike, for example, `LaTeX` [@lamportLATEXDocumentPreparation1994]), works well with version control systems, and can be exported to various document formats, such as HTML websites, a Microsoft Word document, a typeset PDF file (for example, via LaTeX journal templates), or a Powerpoint presentation.
`Markdown` can be used for all sorts of academic documents, ranging from simple sketches of ideas to scientific manuscripts [@R-rticles] and presentations [@revealjs], or even résumés [@vitae].
`R Markdown` extends regular `Markdown` by allowing users to include R code chunks (in fact, arbitrary computer code [@xieMarkdownCookbook2020, Chapter 15, Chapter 15, Other Languages]) into a `Markdown` document.
Upon rendering the document, the code blocks are executed, and their output is dynamically inserted into the document.
This allows the creation of (conditionally) formatted text, statistical results, and figures that are guaranteed to be up-to-date because they are created anew every time the document is rendered to its output format (e.g., presentation slides or a journal article).
@xieMarkdownCookbook2020 provides an extensive yet practical introduction to most features of `R Markdown`.

@@

*Dependency tracking* helps automatically resolve dependencies between project files.
In essence, researchers provide a collection of *computational recipes*.
A computational recipe describes how inputs are processed to deterministically create a specific output in a way that is automatically executable.
The concept of computational recipes is central to our understanding of reproducibility because it enables a unified way to reproduce a project automatically.
Similar to a collection of cooking recipes, we can have multiple products (_targets_) with different ingredients (_requirements_) and different steps of preparation (_recipes_).
In the context of scientific data analysis, targets are typically the final scientific report (e.g., the one to be submitted to a journal) and possibly intermediate results (such as preprocessed data files, simulation results, and analysis results).
A workflow that involves renaming variable names by hand in a graphical spreadsheet application, for example, is therefore incompatible with automated reproducibility.
Another property of a computational recipe is that the same inputs should always result in the same outputs.
For most computer code (given the same software is used), this property is fulfilled.
However, one noteworthy exception is the generation of pseudo-random numbers.
Whenever random numbers are used in a computation, it is only reproducible if the random number generator generates the same numbers.
To ensure identical random numbers, users may fix the state of the random number generated with a so-called seed (e.g. `set.seed()` in R), but they also need to guarantee that the pseudo-random number generator is unchanged [see @Peikert2019].

@@

A version-controlled dynamic document with dependency tracking still relies on external software.
Troubleshooting issues specific to a particular programming language or dependent tool typically require considerable expertise and threaten reproducibility.
*Software management* refers to the act of providing records of, or access to, all software packages and system libraries a project depends on.
One comprehensive approach to software management is containerization.
The central idea is that by
"[..] packaging the key elements of the computational environment needed to run the desired software [makes] the software much easier to use, and the results easier to reproduce [...]" [@silverSoftwareSimplified2017, p. 174].
